---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.path = "man/figures/README-", fig.width = 7, fig.height = 4.5,
  collapse = TRUE, comment = "#>"
)
# Source package functions for rendering before installation
for (f in sort(list.files("R", full.names = TRUE))) source(f)
library(grf)
library(Matrix)
```

# forestBalance

**Forest Kernel Energy Balancing for Causal Inference**

`forestBalance` estimates average treatment effects (ATE) by combining
multivariate random forests with kernel energy balancing. A joint forest model
of covariates, treatment, and outcome defines a proximity kernel that characterizes
the confounding structure and emphasizes similarity of observations in terms of confounding. 
Distributional balancing weights are then obtained via a closed-form
kernel energy distance solution. By construction, these balancing weights aim to balance the joint
distribution of confounders specifically.

The method is described in:

> De, B. and Huling, J. (2025). *Data adaptive covariate balancing for causal effect 
> estimation for high dimensional data.* arXiv:2512.18069.

## Installation

```{r install, eval=FALSE}
# Install from GitHub
devtools::install_github("jaredhuling/forestBalance")
```

## Quick start

```{r quickstart}
# library(forestBalance)

# Simulate observational data with nonlinear confounding (true ATE = 0)
set.seed(42)
dat <- simulate_data(n = 500, p = 10, ate = 0)

# Estimate ATE with forest kernel energy balancing
fit <- forest_balance(dat$X, dat$A, dat$Y)
fit
```

## How it works

The method proceeds in three steps:

1. **Joint forest model:** A `grf::multi_regression_forest` is fit on
   covariates $X$ with the bivariate response $(A, Y)$. Because the forest
   splits on both treatment and outcome, the resulting tree structure captures
   confounding relationships.

2. **Proximity kernel:** The $n \times n$ kernel matrix $K(i,j)$ is defined as
   the proportion of trees where observations $i$ and $j$ share a leaf node.
   This is computed efficiently via a single sparse matrix cross-product.

3. **Kernel energy balancing:** Balancing weights are obtained in closed form
   by solving a linear system derived from the kernel energy distance objective.
   The weights make the treated and control distributions similar with respect
   to the forest-defined similarity measure.

## Detailed example

### Simulating data

`simulate_data()` generates observational data with nonlinear confounding
through a Beta density link:

```{r simulate}
set.seed(123)
dat <- simulate_data(n = 800, p = 10, ate = 0)

# Naive (unweighted) estimate is biased
naive_ate <- mean(dat$Y[dat$A == 1]) - mean(dat$Y[dat$A == 0])
cat("Naive ATE:", round(naive_ate, 4), " (true ATE = 0)\n")
```

### Fitting the model

```{r fit}
fit <- forest_balance(dat$X, dat$A, dat$Y, num.trees = 1000)
```

### Print and summary

`print()` gives a concise overview:

```{r print}
fit
```

`summary()` provides a full covariate balance comparison (unweighted vs
weighted) with flagged imbalances:

```{r summary}
summary(fit)
```

### Balance on nonlinear transformations

Since confounding operates through nonlinear functions of $X_1$ and $X_5$, we
can check balance on transformations of the covariates:

```{r summary-trans}
X <- dat$X
X.nl <- cbind(
  X[,1]^2, X[,2]^2, X[,5]^2,
  X[,1] * X[,2], X[,1] * X[,5],
  dbeta(X[,1], 2, 4), dbeta(X[,5], 2, 4)
)
colnames(X.nl) <- c("X1^2", "X2^2", "X5^2", "X1*X2", "X1*X5",
                     "Beta(X1)", "Beta(X5)")

summary(fit, X.trans = X.nl)
```

### Standalone balance diagnostics

`compute_balance()` can be used independently with any set of weights:

```{r standalone-balance}
# Inverse propensity weights (using true propensity scores)
ipw <- ifelse(dat$A == 1, 1 / dat$propensity, 1 / (1 - dat$propensity))

bal_forest <- compute_balance(dat$X, dat$A, fit$weights)
bal_ipw    <- compute_balance(dat$X, dat$A, ipw)

cat("Forest balance max |SMD|:", round(bal_forest$max_smd, 4), "\n")
cat("IPW balance max |SMD|:   ", round(bal_ipw$max_smd, 4), "\n")
```

### Lower-level interface

For more control, the pipeline can be run step by step:

```{r lower-level}
library(grf)

# 1. Fit the joint forest
forest <- multi_regression_forest(dat$X, scale(cbind(dat$A, dat$Y)),
                                  num.trees = 500)

# 2. Extract leaf node matrix and build kernel
leaf_mat <- get_leaf_node_matrix(forest, dat$X)
K <- leaf_node_kernel(leaf_mat)

cat("Leaf matrix:", nrow(leaf_mat), "obs x", ncol(leaf_mat), "trees\n")
cat("Kernel: ", round(100 * length(K@x) / prod(dim(K)), 1), "% nonzero\n")

# 3. Compute balancing weights
bal <- kernel_balance(dat$A, K)

ate <- weighted.mean(dat$Y[dat$A == 1], bal$weights[dat$A == 1]) -
       weighted.mean(dat$Y[dat$A == 0], bal$weights[dat$A == 0])
cat("ATE estimate:", round(ate, 4), "\n")
```

## Simulation study

A small simulation comparing the forest balance estimator against the naive
(unadjusted) difference in means:

```{r simulation, cache=TRUE}
set.seed(1)
nreps <- 100
results <- matrix(NA, nreps, 2, dimnames = list(NULL, c("Naive", "Forest")))

for (r in seq_len(nreps)) {
  dat <- simulate_data(n = 500, p = 10, ate = 0)
  fit <- forest_balance(dat$X, dat$A, dat$Y, num.trees = 500)
  results[r, "Naive"]  <- mean(dat$Y[dat$A == 1]) - mean(dat$Y[dat$A == 0])
  results[r, "Forest"] <- fit$ate
}
```

```{r sim-results, echo=FALSE}
cat(sprintf("%-10s %8s %8s %8s\n", "", "Bias", "SD", "RMSE"))
cat(strrep("-", 36), "\n")
for (m in colnames(results)) {
  bias <- mean(results[, m])
  sdev <- sd(results[, m])
  rmse <- sqrt(bias^2 + sdev^2)
  cat(sprintf("%-10s %8.4f %8.4f %8.4f\n", m, bias, sdev, rmse))
}
```

```{r sim-plot, echo=FALSE, fig.height=3.5}
par(mar = c(4, 0.5, 2, 0.5))
boxplot(results, horizontal = TRUE, col = c("firebrick", "steelblue"),
        main = "ATE estimates across 100 replications (true ATE = 0)",
        xlab = "Estimated ATE")
abline(v = 0, lty = 2, col = "grey40")
```

## Key functions

| Function | Description |
|---|---|
| `forest_balance()` | High-level: fit forest, build kernel, compute weights, return ATE |
| `simulate_data()` | Simulate observational data with nonlinear confounding |
| `compute_balance()` | Covariate balance diagnostics (SMD, ESS, energy distance) |
| `get_leaf_node_matrix()` | Fast vectorized leaf node extraction from grf forests |
| `leaf_node_kernel()` | Sparse proximity kernel from leaf node matrix |
| `forest_kernel()` | Convenience: forest object to kernel in one call |
| `kernel_balance()` | Closed-form kernel energy balancing weights |
