---
title: "Getting Started with forestBalance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with forestBalance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 6.5, fig.height = 4, collapse = TRUE, comment = "#>"
)
```

## Overview

`forestBalance` estimates average treatment effects (ATE) in observational
studies by combining multivariate random forests with kernel energy balancing.
The key idea is:

1. Fit a random forest that jointly predicts treatment *and* outcome from
   covariates, so that the tree structure captures confounding.
2. Use the forest's leaf co-membership to define a similarity kernel.
3. Obtain balancing weights via a closed-form kernel energy distance solution.

Because the kernel reflects the nonlinear relationships between covariates,
treatment, and outcome, the resulting weights can balance complex confounding
structure that linear methods may miss.

The method is described in:

> De, S. and Huling, J.D. (2025). *Data adaptive covariate balancing for causal
> effect estimation for high dimensional data.* arXiv:2512.18069.

## Setup

```{r packages}
library(forestBalance)
library(WeightIt)
```

## Simulating data

`simulate_data()` generates observational data with nonlinear confounding.  The
propensity score depends on $X_1$ through a Beta density, and the outcome
depends nonlinearly on $X_1$, $X_2$, and $X_5$:

```{r data}
set.seed(123)
dat <- simulate_data(n = 800, p = 10, ate = 0)

c("True ATE"  = dat$ate,
  "Naive ATE" = round(mean(dat$Y[dat$A == 1]) - mean(dat$Y[dat$A == 0]), 4))
```

The naive difference-in-means is badly biased because of confounding.

## Estimating the ATE

### Forest balance

```{r forest-balance}
fit_fb <- forest_balance(dat$X, dat$A, dat$Y, num.trees = 1000)
fit_fb
```

### Entropy balancing (WeightIt)

Entropy balancing finds weights that exactly balance covariate means between
treated and control groups:

```{r ebal}
df <- data.frame(A = dat$A, dat$X)
fit_ebal <- weightit(A ~ ., data = df, method = "ebal")

ate_ebal <- weighted.mean(dat$Y[dat$A == 1], fit_ebal$weights[dat$A == 1]) -
            weighted.mean(dat$Y[dat$A == 0], fit_ebal$weights[dat$A == 0])
```

### Energy balancing (WeightIt)

Energy balancing minimizes the energy distance between the weighted treated and
control covariate distributions:

```{r energy}
fit_energy <- weightit(A ~ ., data = df, method = "energy")

ate_energy <- weighted.mean(dat$Y[dat$A == 1], fit_energy$weights[dat$A == 1]) -
              weighted.mean(dat$Y[dat$A == 0], fit_energy$weights[dat$A == 0])
```

### Comparison

```{r ate-comparison, echo=FALSE}
ests <- data.frame(
  Method = c("Naive", "Entropy balancing", "Energy balancing",
             "Forest balance", "Truth"),
  ATE = round(c(mean(dat$Y[dat$A == 1]) - mean(dat$Y[dat$A == 0]),
                ate_ebal, ate_energy, fit_fb$ate, dat$ate), 4),
  stringsAsFactors = FALSE
)
knitr::kable(ests, caption = "Single-replication ATE estimates.")
```

## Covariate balance

`summary()` shows the full balance comparison.  Let's also check balance on
nonlinear transformations of the covariates, since the confounding operates
through the Beta density:

```{r balance}
X <- dat$X
X.nl <- cbind(
  X[, 1]^2, X[, 2]^2, X[, 5]^2,
  X[, 1] * X[, 2], X[, 1] * X[, 5],
  dbeta(X[, 1], 2, 4), dbeta(X[, 5], 2, 4)
)
colnames(X.nl) <- c("X1^2", "X2^2", "X5^2", "X1*X2", "X1*X5",
                     "Beta(X1)", "Beta(X5)")

summary(fit_fb, X.trans = X.nl)
```

We can use `compute_balance()` to compare all methods on the same terms:

```{r balance-comparison}
bal_fb     <- compute_balance(dat$X, dat$A, fit_fb$weights, X.trans = X.nl)
bal_ebal   <- compute_balance(dat$X, dat$A, fit_ebal$weights, X.trans = X.nl)
bal_energy <- compute_balance(dat$X, dat$A, fit_energy$weights, X.trans = X.nl)
bal_unwtd  <- compute_balance(dat$X, dat$A, rep(1, dat$n), X.trans = X.nl)
```

```{r balance-table, echo=FALSE}
btbl <- data.frame(
  Method = c("Unweighted", "Entropy balancing", "Energy balancing", "Forest balance"),
  `Max SMD (linear)` = round(c(bal_unwtd$max_smd, bal_ebal$max_smd,
                                bal_energy$max_smd, bal_fb$max_smd), 4),
  `Max SMD (nonlinear)` = round(c(bal_unwtd$max_smd_trans, bal_ebal$max_smd_trans,
                                   bal_energy$max_smd_trans, bal_fb$max_smd_trans), 4),
  `ESS treated` = paste0(round(c(1, bal_ebal$ess_treated,
                                  bal_energy$ess_treated, bal_fb$ess_treated) * 100), "%"),
  `ESS control` = paste0(round(c(1, bal_ebal$ess_control,
                                  bal_energy$ess_control, bal_fb$ess_control) * 100), "%"),
  check.names = FALSE
)
knitr::kable(btbl, caption = "Balance diagnostics across methods.")
```

All three weighting methods reduce the linear covariate imbalance.  Both energy balancing the forest 
balance approach also reduce imbalance on *nonlinear* functions of the covariates, as they both
aim to balance the joint distributions of covariates, however, forest balancing over-emphasizes full 
distributional balance of confounders specifically rather than all possible covariates.

## Simulation study

A simulation study with 100 replications at two sample sizes ($n = 500$ and
$n = 1{,}000$) demonstrates how the methods compare and how performance changes
with $n$:

```{r simulation, cache=TRUE}
run_sim <- function(n, nreps = 50, num.trees = 500, seed = 1) {
  set.seed(seed)
  methods <- c("Naive", "Entropy", "Energy", "Forest")
  results <- matrix(NA, nreps, length(methods), dimnames = list(NULL, methods))

  for (r in seq_len(nreps)) {
    dat <- simulate_data(n = n, p = 10, ate = 0)
    df  <- data.frame(A = dat$A, dat$X)

    fit_fb <- forest_balance(dat$X, dat$A, dat$Y, num.trees = num.trees)

    w_ebal   <- tryCatch(weightit(A ~ ., data = df, method = "ebal")$weights,
                          error = function(e) rep(1, dat$n))
    w_energy <- tryCatch(weightit(A ~ ., data = df, method = "energy")$weights,
                          error = function(e) rep(1, dat$n))

    results[r, "Naive"]   <- mean(dat$Y[dat$A == 1]) - mean(dat$Y[dat$A == 0])
    results[r, "Entropy"] <- weighted.mean(dat$Y[dat$A == 1], w_ebal[dat$A == 1]) -
                              weighted.mean(dat$Y[dat$A == 0], w_ebal[dat$A == 0])
    results[r, "Energy"]  <- weighted.mean(dat$Y[dat$A == 1], w_energy[dat$A == 1]) -
                              weighted.mean(dat$Y[dat$A == 0], w_energy[dat$A == 0])
    results[r, "Forest"]  <- fit_fb$ate
  }
  results
}

res_500  <- run_sim(n = 500,  seed = 1)
res_1000 <- run_sim(n = 1000, seed = 1)
```

```{r sim-table, echo=FALSE}
make_summary <- function(results, n_label) {
  data.frame(
    n      = n_label,
    Method = colnames(results),
    Bias   = round(colMeans(results), 4),
    SD     = round(apply(results, 2, sd), 4),
    RMSE   = round(sqrt(colMeans(results)^2 + apply(results, 2, var)), 4)
  )
}
stbl <- rbind(make_summary(res_500, 500), make_summary(res_1000, 1000))
knitr::kable(stbl, row.names = FALSE,
             caption = "Simulation results (50 reps, true ATE = 0).")
```

```{r sim-plot, echo=FALSE, fig.height=4.5, fig.width=6.5}
library(ggplot2)

# Combine both sample sizes into long format
to_long <- function(mat, n_label) {
  data.frame(
    ATE = as.vector(mat),
    Method = rep(colnames(mat), each = nrow(mat)),
    n = paste0("n = ", format(n_label, big.mark = ","))
  )
}
plot_df <- rbind(to_long(res_500, 500), to_long(res_1000, 1000))
plot_df$Method <- factor(plot_df$Method,
                         levels = c("Naive", "Entropy", "Energy", "Forest"))
plot_df$n <- factor(plot_df$n, levels = c("n = 500", "n = 1,000"))

ggplot(plot_df, aes(x = ATE, y = Method, fill = Method)) +
  geom_boxplot(outlier.size = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  scale_fill_manual(values = c(Naive = "firebrick", Entropy = "darkorange",
                                Energy = "goldenrod3", Forest = "steelblue")) +
  facet_wrap(~ n, ncol = 1, scales = "free_y") +
  labs(x = "Estimated ATE", y = NULL) +
  theme_bw(base_size = 12) +
  theme(legend.position = "none",
        strip.text = element_text(face = "bold"))
```

All weighting methods improve over the naive estimator.  At $n = 500$, entropy
balancing (which only targets linear covariate means) retains substantial bias,
while energy balancing and forest balance both reduce bias more effectively.
At $n = 1{,}000$, all methods improve, but forest balance continues to show
the lowest bias and competitive RMSE, reflecting the advantage of its
confounding-targeted kernel.

## Step-by-step interface

For more control, the pipeline can be run in stages:

```{r stepwise}
library(grf)

dat <- simulate_data(n = 500, p = 10)

# 1. Fit the joint forest
forest <- multi_regression_forest(dat$X, scale(cbind(dat$A, dat$Y)),
                                  num.trees = 500)

# 2. Extract leaf node matrix (n x B)
leaf_mat <- get_leaf_node_matrix(forest, dat$X)
dim(leaf_mat)

# 3. Build sparse proximity kernel
K <- leaf_node_kernel(leaf_mat)
c("kernel % nonzero" = round(100 * length(K@x) / prod(dim(K)), 1))

# 4. Compute balancing weights
bal <- kernel_balance(dat$A, K)

# 5. Estimate ATE
weighted.mean(dat$Y[dat$A == 1], bal$weights[dat$A == 1]) -
  weighted.mean(dat$Y[dat$A == 0], bal$weights[dat$A == 0])
```
