---
title: "Performance and Scalability"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Performance and Scalability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 6.5, fig.height = 4, collapse = TRUE, comment = "#>"
)
```

## Overview

This vignette benchmarks the `forestBalance` pipeline---leaf node extraction,
kernel construction, and weight computation---to characterize how speed, memory,
and kernel sparsity scale with sample size ($n$) and number of trees ($B$).

The two computationally critical components are:

- **`get_leaf_node_matrix()`**: traverses all $B$ trees for $n$ observations.
  Implemented in C++ via Rcpp.
- **`leaf_node_kernel()`**: constructs a sparse indicator matrix and computes a
  single sparse `tcrossprod`.

We benchmark up to $n = 25{,}000$ to demonstrate feasibility at scales where a
dense $n \times n$ kernel matrix would require 4.7 GB.

```{r packages}
library(forestBalance)
library(grf)
library(Matrix)
```

## Benchmark

```{r benchmark, cache=TRUE}
n_vals    <- c(500, 1000, 2500, 5000, 10000, 25000)
tree_vals <- c(500, 1000, 2000)
p <- 10

results <- data.frame()

for (nn in n_vals) {
  for (B in tree_vals) {
    set.seed(123)
    dat <- simulate_data(n = nn, p = p)

    # Fit forest
    t_fit <- system.time({
      fit <- multi_regression_forest(
        dat$X, scale(cbind(dat$A, dat$Y)),
        num.trees = B, min.node.size = 10
      )
    })["elapsed"]

    # Leaf node extraction
    t_leaf <- system.time({
      leaf_mat <- get_leaf_node_matrix(fit, dat$X)
    })["elapsed"]

    # Kernel construction
    t_kern <- system.time({
      K <- leaf_node_kernel(leaf_mat)
    })["elapsed"]

    # Kernel balance
    t_bal <- system.time({
      w <- kernel_balance(dat$A, K)
    })["elapsed"]

    # Stats
    nnz <- length(K@x)
    pct_nz <- 100 * nnz / as.numeric(nn)^2
    mem_sparse <- as.numeric(object.size(K)) / 1e6
    mem_dense  <- 8 * as.numeric(nn)^2 / 1e6

    results <- rbind(results, data.frame(
      n = nn, trees = B,
      t_fit = t_fit, t_leaf = t_leaf, t_kern = t_kern, t_bal = t_bal,
      t_total = t_fit + t_leaf + t_kern + t_bal,
      nnz_M = nnz / 1e6, pct_nz = pct_nz,
      sparse_MB = mem_sparse, dense_MB = mem_dense
    ))
  }
}
```

## Timing

The table below shows wall-clock time for each pipeline stage.  "Total" includes
forest fitting, leaf extraction, kernel construction, and weight computation.

```{r timing-table, echo=FALSE}
tbl <- results
tbl$t_fit  <- round(tbl$t_fit, 1)
tbl$t_leaf <- round(tbl$t_leaf, 2)
tbl$t_kern <- round(tbl$t_kern, 2)
tbl$t_bal  <- round(tbl$t_bal, 2)
tbl$t_total <- round(tbl$t_total, 1)
disp <- tbl[, c("n", "trees", "t_fit", "t_leaf", "t_kern", "t_bal", "t_total")]
names(disp) <- c("n", "Trees", "Fit (s)", "Leaf (s)", "Kernel (s)",
                  "Balance (s)", "Total (s)")
knitr::kable(disp, row.names = FALSE, align = "rrrrrrr",
             caption = "Pipeline timing by stage.")
```

Leaf node extraction and kernel construction both scale approximately linearly
in $n$ and $B$.  The forest fitting itself (handled by `grf`) is typically the
fastest step.

```{r timing-plot, echo=FALSE, fig.height=4}
cols <- c("500" = "#E69F00", "1000" = "#56B4E9", "2000" = "#009E73")

par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))

# Total time vs n
plot(NULL, xlim = range(results$n), ylim = range(results$t_total),
     xlab = "n", ylab = "Total time (s)", log = "xy",
     main = "Total time vs n")
for (tr in unique(results$trees)) {
  sub <- results[results$trees == tr, ]
  lines(sub$n, sub$t_total, type = "b", pch = 16,
        col = cols[as.character(tr)], lwd = 2)
}
legend("topleft", legend = paste(names(cols), "trees"),
       col = cols, pch = 16, lty = 1, lwd = 2, bty = "n", cex = 0.85)

# Time breakdown at B=1000
sub1k <- results[results$trees == 1000, ]
bardata <- t(as.matrix(sub1k[, c("t_fit", "t_leaf", "t_kern", "t_bal")]))
colnames(bardata) <- format(sub1k$n, big.mark = ",")
barplot(bardata, col = c("grey70", "#56B4E9", "#009E73", "#CC79A7"),
        xlab = "n", ylab = "Time (s)", main = "Breakdown (1000 trees)",
        legend.text = c("Forest fit", "Leaf extract", "Kernel build", "Balance"),
        args.legend = list(cex = 0.7, bty = "n"))
```

## Kernel sparsity

The proximity kernel becomes sparser as $n$ increases: with more observations,
fewer pairs share a leaf in any given tree.  At $n = 25{,}000$ with 2,000 trees,
only about 5% of entries are nonzero.

```{r sparsity-table, echo=FALSE}
stbl <- results[, c("n", "trees", "pct_nz", "nnz_M")]
stbl$pct_nz <- paste0(round(stbl$pct_nz, 1), "%")
stbl$nnz_M  <- round(stbl$nnz_M, 1)
names(stbl) <- c("n", "Trees", "% nonzero", "Nonzeros (M)")
knitr::kable(stbl, row.names = FALSE, align = "rrlr",
             caption = "Kernel sparsity across configurations.")
```

```{r sparsity-plot, echo=FALSE, fig.height=3.5}
par(mar = c(4, 4, 2, 1))
plot(NULL, xlim = range(results$n), ylim = c(0, max(results$pct_nz)),
     xlab = "n", ylab = "% nonzero", main = "Kernel density vs n")
for (tr in unique(results$trees)) {
  sub <- results[results$trees == tr, ]
  lines(sub$n, sub$pct_nz, type = "b", pch = 16,
        col = cols[as.character(tr)], lwd = 2)
}
legend("topright", legend = paste(names(cols), "trees"),
       col = cols, pch = 16, lty = 1, lwd = 2, bty = "n")
```

## Memory: sparse vs dense

A dense $n \times n$ double matrix requires $8n^2$ bytes.  At $n = 25{,}000$
this is **4.7 GB**---well beyond typical RAM budgets for a single matrix.  The
sparse representation exploits the kernel's low density and grows much more
slowly.

```{r memory-table, echo=FALSE}
mtbl <- results[, c("n", "trees", "sparse_MB", "dense_MB")]
mtbl$ratio <- paste0(round(mtbl$sparse_MB / mtbl$dense_MB * 100, 1), "%")
mtbl$sparse_MB <- round(mtbl$sparse_MB, 1)
mtbl$dense_MB  <- format(round(mtbl$dense_MB, 0), big.mark = ",")
names(mtbl) <- c("n", "Trees", "Sparse (MB)", "Dense (MB)", "Sparse / Dense")
knitr::kable(mtbl, row.names = FALSE, align = "rrrrl",
             caption = "Kernel memory: sparse (actual) vs dense (theoretical).")
```

```{r memory-plot, echo=FALSE, fig.height=4}
par(mar = c(4, 4.5, 2, 1))
plot(NULL, xlim = range(results$n),
     ylim = c(min(results$sparse_MB), max(results$dense_MB)),
     xlab = "n", ylab = "Memory (MB)", log = "xy",
     main = "Kernel memory: sparse vs dense")

# Dense reference
dns <- results[results$trees == results$trees[1], ]
lines(dns$n, dns$dense_MB, type = "b", pch = 4, col = "firebrick",
      lwd = 2, lty = 2)

for (tr in unique(results$trees)) {
  sub <- results[results$trees == tr, ]
  lines(sub$n, sub$sparse_MB, type = "b", pch = 16,
        col = cols[as.character(tr)], lwd = 2)
}
legend("topleft",
       legend = c("Dense (theoretical)", paste(names(cols), "trees (sparse)")),
       col = c("firebrick", cols), pch = c(4, 16, 16, 16),
       lty = c(2, 1, 1, 1), lwd = 2, bty = "n", cex = 0.8)
```

At $n = 25{,}000$ the sparse kernel uses about **6--8%** of what a dense matrix
would require, making problems at this scale tractable on a standard laptop.

## Summary

```{r summary-table, echo=FALSE}
big <- results[results$n == max(results$n) & results$trees == max(results$trees), ]
cat(sprintf("Largest configuration tested: n = %s, B = %s\n",
            format(big$n, big.mark = ","), format(big$trees, big.mark = ",")))
cat(sprintf("  Total pipeline time: %.0f seconds\n", big$t_total))
cat(sprintf("  Kernel sparsity:     %.1f%% nonzero\n", big$pct_nz))
cat(sprintf("  Sparse memory:       %.0f MB (vs %.0f MB dense)\n",
            big$sparse_MB, big$dense_MB))
```

### Key takeaways:

- **Leaf extraction** (C++) and **kernel construction** (sparse `tcrossprod`)
  each scale approximately linearly in both $n$ and $B$.
- The **kernel is naturally sparse** and becomes sparser as $n$ grows, making
  the sparse representation increasingly advantageous.
- At $n = 25{,}000$, the full pipeline (forest fit + kernel + weights) runs in
  under two minutes, using a fraction of the memory a dense approach would need.
