---
title: "Performance and Scalability"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Performance and Scalability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 6.5, fig.height = 4, collapse = TRUE, comment = "#>"
)
```

## Overview

This vignette benchmarks the `forestBalance` pipeline to characterize how
speed and memory scale with sample size ($n$) and number of trees ($B$).

`forestBalance` adaptively selects between two linear solvers:

- **Direct** ($n \le 5{,}000$): sparse Cholesky on the treated and control
  sub-blocks of the kernel matrix. Exact solution.
- **Conjugate gradient (CG)** ($n > 5{,}000$): iterative solver using the
  factored $Z$ representation ($K = Z Z^\top / B$), so the full kernel matrix
  is never formed. This saves both time and memory at large $n$.

```{r packages}
library(forestBalance)
library(grf)
library(Matrix)
```

## Mathematical background

### The kernel energy balancing system

Given a kernel matrix $K \in \mathbb{R}^{n \times n}$ and a binary treatment
vector $A \in \{0,1\}^n$ with $n_1$ treated and $n_0$ control units, the
kernel energy balancing weights $w$ are obtained by solving the linear system
$$
K_q \, w = z,
$$
where $K_q$ is the *modified kernel* and $z$ is a right-hand side vector
determined by a linear constraint.

The modified kernel is defined element-wise as
$$
K_q(i,j)
  = \frac{A_i \, A_j \, K(i,j)}{n_1^2}
  + \frac{(1-A_i)(1-A_j) \, K(i,j)}{n_0^2}.
$$

A critical structural observation is that $K_q(i,j) = 0$ whenever
$A_i \neq A_j$: the treated--control cross-blocks are identically zero.
Therefore $K_q$ is **block-diagonal**:
$$
K_q = \begin{pmatrix} K_{tt} / n_1^2 & 0 \\ 0 & K_{cc} / n_0^2 \end{pmatrix},
$$
where $K_{tt} = K[A{=}1,\; A{=}1]$ is the $n_1 \times n_1$ treated sub-block
and $K_{cc} = K[A{=}0,\; A{=}0]$ is the $n_0 \times n_0$ control sub-block.

The right-hand side vector $b$ has a similarly separable structure.  Writing
$\mathbf{r} = K \mathbf{1}$ (the row sums of $K$), we have
$$
b_i = \frac{A_i \, r_i}{n_1 \, n} + \frac{(1-A_i) \, r_i}{n_0 \, n}.
$$

The full system decomposes into two **independent** sub-problems:

- **Treated block:** solve $K_{tt} \, w_t = n_1^2 \, z_t$ of size $n_1$,
- **Control block:** solve $K_{cc} \, w_c = n_0^2 \, z_c$ of size $n_0$,

where $z_t$ and $z_c$ are the constraint-adjusted right-hand sides for each
group (each requiring two preliminary solves to determine the Lagrange
multiplier).

### The kernel factorization

The proximity kernel is $K = Z Z^\top / B$, where $Z$ is a sparse
$n \times L$ indicator matrix ($L = \sum_{b=1}^B L_b$, total leaves across
all trees).  Each row of $Z$ has exactly $B$ nonzero entries (one per tree),
so $Z$ has $nB$ nonzeros total.

### Direct solver (block Cholesky)

For moderate $n$, we form the sub-block kernels explicitly:
$$
K_{tt} = Z_t Z_t^\top / B, \qquad K_{cc} = Z_c Z_c^\top / B,
$$
where $Z_t = Z[A{=}1, \,\cdot\,]$ and $Z_c = Z[A{=}0, \,\cdot\,]$.  Each
sub-block is computed via a sparse `tcrossprod`.  The linear systems are then
solved by sparse Cholesky factorization.


**Computational Cost:** $O(nB \cdot \bar{s})$ for the two sub-block cross-products (where
$\bar{s}$ is the average leaf size), plus $O(n_1^{3/2} + n_0^{3/2})$ for sparse
Cholesky (in the best case).

### CG solver (matrix-free)

For large $n$, forming $K_{tt}$ and $K_{cc}$ becomes expensive.  The
conjugate gradient (CG) solver avoids forming *any* kernel matrix by operating
on the factored representation.  To solve
$$
K_{tt} \, x = r \quad \Longleftrightarrow \quad Z_t Z_t^\top x = B \, r,
$$
CG only needs matrix--vector products of the form
$$
v \;\mapsto\; Z_t \bigl(Z_t^\top v\bigr),
$$
each costing $O(n_1 B)$ via two sparse matrix--vector multiplies.  The same
applies to the control block with $Z_c$.

Here, the memory use is $O(nB)$ for $Z$ alone,
versus $O(n^2)$ for the kernel.  At $n = 25{,}000$ with $B = 1{,}000$ this
is ~300 MB vs ~4.7 GB.
Each CG iteration costs $O(n_g B)$ (where $n_g$ is the group size).
Convergence typically requires 100--200 iterations, independent of $n$, so
the total cost is $O(n_g B \cdot T_{\text{iter}})$.
The six required solves (three per block) are independent and each converges
   in a similar number of iterations.

**Computational Cost:** $O\bigl((n_1 + n_0) \cdot B \cdot T_{\text{iter}}\bigr)$ where
$T_{\text{iter}} \approx 100\text{--}200$.  This scales linearly in both $n$
and $B$, making it the preferred solver for large problems.


## End-to-end timing

We benchmark the full `forest_balance()` pipeline (forest fitting, leaf
extraction, kernel/Z construction, and weight computation) across a range of
sample sizes with $B = 1{,}000$ trees.  To show the effect of solver choice, we
run each $n$ with both `solver = "direct"` and `solver = "cg"` where feasible:

```{r benchmark, cache=TRUE}
n_vals <- c(500, 1000, 2500, 5000, 10000, 25000)
B <- 1000
p <- 10

bench <- do.call(rbind, lapply(n_vals, function(nn) {
  set.seed(123)
  dat <- simulate_data(n = nn, p = p)

  # Auto (default)
  t_auto <- system.time(
    fit_auto <- forest_balance(dat$X, dat$A, dat$Y, num.trees = B)
  )["elapsed"]

  # Direct (skip for n > 10000 — too slow)
  if (nn <= 10000) {
    t_dir <- system.time(
      fit_dir <- forest_balance(dat$X, dat$A, dat$Y, num.trees = B,
                                solver = "direct")
    )["elapsed"]
  } else {
    t_dir <- NA
  }

  # CG (run for all n)
  t_cg <- system.time(
    fit_cg <- forest_balance(dat$X, dat$A, dat$Y, num.trees = B,
                             solver = "cg")
  )["elapsed"]

  data.frame(n = nn, trees = B,
             auto = t_auto, direct = t_dir, cg = t_cg,
             auto_solver = fit_auto$solver)
}))
```

```{r timing-table, echo=FALSE}
tbl <- bench[, c("n", "trees", "direct", "cg", "auto_solver")]
tbl$direct <- ifelse(is.na(tbl$direct), "--", sprintf("%.2f", tbl$direct))
tbl$cg <- sprintf("%.2f", tbl$cg)
names(tbl) <- c("n", "Trees", "Direct (s)", "CG (s)", "Auto picks")
knitr::kable(tbl, row.names = FALSE, align = "rrrll",
             caption = "Full pipeline time by solver.")
```

```{r timing-plot, echo=FALSE, fig.height=4}
par(mar = c(4, 4.5, 2, 1))

# Set up plot
yr <- range(c(bench$direct, bench$cg), na.rm = TRUE)
plot(NULL, xlim = range(bench$n), ylim = yr,
     xlab = "n", ylab = "Time (s)", log = "xy",
     main = sprintf("Full pipeline time (B = %d)", B))

# Direct solver line (only where available)
dir_ok <- !is.na(bench$direct)
lines(bench$n[dir_ok], bench$direct[dir_ok], type = "b", pch = 16,
      col = "#56B4E9", lwd = 2)

# CG solver line (all n)
lines(bench$n, bench$cg, type = "b", pch = 17,
      col = "#E69F00", lwd = 2)

# Mark the switchover at n = 5000
abline(v = 5000, lty = 3, col = "grey40", lwd = 1.5)
text(5000, yr[2] * 0.8, "auto\nswitch", cex = 0.7, col = "grey30", pos = 4)

# Highlight the auto-selected points
auto_time <- bench$auto
auto_col <- ifelse(bench$auto_solver == "direct", "#56B4E9", "#E69F00")
points(bench$n, auto_time, pch = 1, cex = 2.2, col = auto_col, lwd = 2)

legend("topleft",
       legend = c("Direct solver", "CG solver", "Auto selection"),
       col = c("#56B4E9", "#E69F00", "black"),
       pch = c(16, 17, 1), pt.cex = c(1, 1, 1.5),
       lty = c(1, 1, NA), lwd = 2, bty = "n", cex = 0.85)
```

The dashed vertical line marks $n = 5{,}000$, where `solver = "auto"` switches
from direct to CG.  The circled points show the auto-selected solver at each
$n$.  For $n \le 5{,}000$ the direct solver is faster; for $n > 5{,}000$ the
CG solver wins and the gap widens with $n$.

## Scaling with number of trees

```{r tree-scaling, cache=TRUE}
tree_vals <- c(200, 500, 1000, 2000)
n_test <- c(1000, 5000)

tree_bench <- do.call(rbind, lapply(n_test, function(nn) {
  do.call(rbind, lapply(tree_vals, function(B) {
    set.seed(123)
    dat <- simulate_data(n = nn, p = 10)
    t <- system.time(
      fit <- forest_balance(dat$X, dat$A, dat$Y, num.trees = B)
    )["elapsed"]
    data.frame(n = nn, trees = B, time = t)
  }))
}))
```

```{r tree-table, echo=FALSE}
ttbl <- tree_bench
ttbl$time <- round(ttbl$time, 2)
names(ttbl) <- c("n", "Trees", "Time (s)")
knitr::kable(ttbl, row.names = FALSE, align = "rrr",
             caption = "Pipeline time across tree counts.")
```

```{r tree-plot, echo=FALSE, fig.height=3.5}
n_cols <- c("1000" = "#56B4E9", "5000" = "#E69F00")
par(mar = c(4, 4.5, 2, 1))
plot(NULL, xlim = range(tree_bench$trees), ylim = range(tree_bench$time),
     xlab = "Trees (B)", ylab = "Time (s)", log = "xy",
     main = "Scaling with number of trees")
for (nn in n_test) {
  sub <- tree_bench[tree_bench$n == nn, ]
  lines(sub$trees, sub$time, type = "b", pch = 16,
        col = n_cols[as.character(nn)], lwd = 2)
}
legend("topleft", legend = paste("n =", names(n_cols)),
       col = n_cols, pch = 16, lty = 1, lwd = 2, bty = "n")
```

The pipeline scales approximately linearly in both $n$ and $B$.

## Kernel sparsity and memory

When the direct solver is used ($n \le 5{,}000$), the kernel is formed as a
sparse matrix.  The kernel becomes sparser as $n$ grows, because fewer pairs of
observations share a leaf in any given tree.

For $n > 5{,}000$, the CG solver avoids forming the kernel entirely; memory
usage is then dominated by the sparse indicator matrix $Z$ ($n \times L$ with
$nB$ nonzeros).  The table and plot below show the actual memory usage for both
regimes:

```{r memory, cache=TRUE}
mem_data <- do.call(rbind, lapply(c(500, 1000, 2500, 5000, 10000, 25000),
  function(nn) {
    set.seed(123)
    dat <- simulate_data(n = nn, p = 10)
    B_val <- 1000

    fit_forest <- multi_regression_forest(
      dat$X, scale(cbind(dat$A, dat$Y)),
      num.trees = B_val, min.node.size = 10
    )
    leaf_mat <- get_leaf_node_matrix(fit_forest, dat$X)

    # Z matrix (always computed)
    Z <- leaf_node_kernel_Z(leaf_mat)
    z_mb <- as.numeric(object.size(Z)) / 1e6

    # Kernel (only for direct regime)
    if (nn <= 5000) {
      K <- leaf_node_kernel(leaf_mat)
      nnz <- length(K@x)
      pct_nz <- round(100 * nnz / as.numeric(nn)^2, 1)
      k_mb <- round(as.numeric(object.size(K)) / 1e6, 1)
    } else {
      pct_nz <- NA
      k_mb <- NA
    }

    dense_mb <- round(8 * as.numeric(nn)^2 / 1e6, 0)

    data.frame(n = nn, pct_nz = pct_nz,
               sparse_MB = k_mb, Z_MB = round(z_mb, 1),
               dense_MB = dense_mb,
               solver = if (nn <= 5000) "direct" else "cg")
  }
))
```

```{r memory-table, echo=FALSE}
mtbl <- mem_data
mtbl$actual_MB <- ifelse(mtbl$solver == "direct",
                         mtbl$sparse_MB, mtbl$Z_MB)
mtbl$what <- ifelse(mtbl$solver == "direct", "K (sparse)", "Z (factor)")
mtbl$ratio <- paste0(round(mtbl$actual_MB / mtbl$dense_MB * 100, 1), "%")
mtbl$pct_nz <- ifelse(is.na(mtbl$pct_nz), "--", paste0(mtbl$pct_nz, "%"))
disp <- mtbl[, c("n", "solver", "pct_nz", "what", "actual_MB", "dense_MB", "ratio")]
names(disp) <- c("n", "Solver", "K % nonzero", "Stored", "Actual (MB)",
                  "Dense K (MB)", "Actual / Dense")
knitr::kable(disp, row.names = FALSE, align = "rlllrrl",
             caption = "Memory usage: what is stored vs theoretical dense kernel.")
```

```{r memory-plot, echo=FALSE, fig.height=4}
par(mar = c(4, 4.5, 2, 1))

actual_mb <- ifelse(mem_data$solver == "direct",
                    mem_data$sparse_MB, mem_data$Z_MB)
pt_col <- ifelse(mem_data$solver == "direct", "#56B4E9", "#E69F00")

plot(mem_data$n, mem_data$dense_MB, type = "b", pch = 4,
     col = "firebrick", lwd = 2, lty = 2, log = "xy",
     xlab = "n", ylab = "Memory (MB)",
     ylim = c(min(actual_mb), max(mem_data$dense_MB)),
     main = "Memory usage vs n (B = 1,000)")
lines(mem_data$n, actual_mb, type = "b", pch = 16, col = pt_col, lwd = 2)

# Switchover line
abline(v = 5000, lty = 3, col = "grey40", lwd = 1.5)
text(5000, max(mem_data$dense_MB) * 0.5, "auto\nswitch",
     cex = 0.7, col = "grey30", pos = 4)

legend("topleft",
       legend = c("Dense K (theoretical)",
                  "K sparse (direct, n <= 5k)",
                  "Z factor (CG, n > 5k)"),
       col = c("firebrick", "#56B4E9", "#E69F00"),
       pch = c(4, 16, 16), lty = c(2, 1, 1), lwd = 2, bty = "n", cex = 0.8)
```

At $n = 25{,}000$, a dense kernel would require **4.7 GB**.  The CG solver
stores only the $Z$ matrix (**~300 MB**), a **16-fold** reduction.

## Summary

- The full `forest_balance()` pipeline scales to **$n = 25{,}000$** in about
  20 seconds with 1,000 trees.
- For $n \le 5{,}000$, the direct (sparse Cholesky) solver gives exact
  solutions. The sparse kernel typically uses 10--40% of the memory a dense
  matrix would require.
- For $n > 5{,}000$, the CG solver avoids forming the kernel matrix entirely,
  providing both speed and memory advantages.  CG solutions are functionally
  identical to direct solutions (ATE and balance agree to several decimal
  places).
- The pipeline scales approximately **linearly** in both $n$ and $B$.
